---
title: 'Assignment4: shiny'
author: "jharner"
date: "April 24, 2015"
output: html_document
---

Consider the variable weight (`wt`) from the `mtcars` data set which is in the builtin `datasets` package. 
```{r}
data(mtcars)
```

Use shiny to construct a histogram of `wt` in which the number of bins is controlled by a slider. Organize and label the plot appropriately.

You will need to construct two files: `ui.R` for the user interface and `server.R` for the server-side code. Explain how the app is run.

Include the code here for reference.
```{r}
library(shiny)
data(mtcars)
shinyUI(fluidPage(
    titlePanel("DSCI 504: Assignment 4"),
    sidebarLayout(
        sidebarPanel(
            sliderInput("n.bins", label = h4("Number of Bins"), min = 2, 
                        max = nrow(mtcars), value = 5, step = 1)),
        mainPanel(plotOutput("hist")))))
```

The user interface code sets up the page using a sidebar layout. The
Sidebar contains a slider for selecting the number of bins, and the
main panel draws the plot generated by the server.

```{r}
data(mtcars)
shinyServer(
    function(input, output){
        output$hist <- renderPlot({
            n.bins <- input$n.bins
            main.title <- paste("Histogram of Vehicle Weight with", n.bins, "Bins")
            hist(mtcars$wt, breaks = n.bins - 1, col = "grey20", border = "grey50",
                 main = main.title, xlab = "Weight")
        })
})
```

The server first loads the data set and initializes the server
session. From here, we create the plot output which extracts the
number of bins from the user interface and draws the histogram.

Note that the number of breaks is not strictly enforced by `hist()`,
as per the documentation:

> ... the number [of bins] is a suggestion only; the  breakpoints will
> be set to ‘pretty’ values.  If ‘breaks’ is a  function, the ‘x’
> vector is supplied to it as the only  argument.

