---
title: 'Chap3: Customizing Traditional Graphics'
author: "jharner"
date: "January 26, 2015"
output: html_document
---

This chapter describes low-level traditional functions that are useful
for controlling the fine details of a plot and for adding further
output to a plot. 

## 3.1 The traditional graphics model in more detail

### 3.1.1 Plotting regions

Every page is split up into three main regions:  

* the **outer margins**  
* the **current figure region**  
* the **current plot region**   

See Figure 3.1 and Figure 3.2 for details.

The region obtained by removing the outer margins from the device is
called the **inner region**. When there is only one figure, this
usually corresponds to the **figure region**. 

The area outside the plot region, but inside the figure region is
referred to as the **figure margins**. A typical high-level function
draws data symbols and lines within the plot region and axes and
labels in the figure margins or outer margins. 

The size and location of the different regions are controlled either
via the `par()` function, or using special functions. 

#### Coordinate systems

The coordinate system in the plot region, referred to as **user
coordinates**, corresponds to the range of values on the axes of the
plot (Figure 3.3). The drawing of data symbols, lines, and text in the
plot region occurs relative to this user coordinate system. 

The coordinate systems in the figure margins are a combination of x-
or y-ranges (like user coordinates) and lines of text away from the
boundary of the plot region (Figure 3.4). **Normalized coordinate
systems** available for the figure margins in which the x- and
y-ranges are replaced with a range from 0 to 1. 

The outer margins have similar sets of coordinate systems, but
locations along the boundary of the inner region can only be specified
in normalized coordinates (always relative to the extent of the
complete outer margin) (Figure 3.5). 

### 3.1.2 The traditional graphics state

The traditional graphics system maintains a **graphics state** for
each graphics device and, when drawing occurs, this state is consulted
to determine where output should be drawn, what color to use, what
fonts to use, etc. 

The graphics state consists of a large number of settings to describe:  

* the size and placement of the plot regions and coordinate systems;   
* the general appearance of graphical output (e.g., the colors and line types that are used to draw lines and the fonts that are used to draw text);  
* aspects of the output device (e.g., the physical size of the device
and the current clipping region).   

See Table 3.1 to 3.3.

The main function used to access the graphics state is the
`par()`function. Simply typing `par()` will result in a complete
listing of the current graphics state. A specific state setting can be
queried by supplying specific setting names as arguments to `par()`. 
```{r}
# par()
par(c("col", "lty"))
```

The par() function can be used to modify traditional graphics state
settings by specifying a value via an argument with the appropriate
setting name. 
```{r}
par(col="red", lty="dashed")
par(col="black", lty="solid") # reset
```

Modifying traditional graphics state settings via `par()` has a
persistent effect. Settings specified in this way will hold until a
different setting is specified. Settings may also be *temporarily*
modified by specifying a new value in a call to a graphics function
such as `plot()` or `lines()`. 
```{r}
y <- rnorm(20)
par(lty="dashed")
plot(y, type="l") # line is dashed
plot(y, type="l", lty="solid") # line is solid
plot(y, type="l") # line is dashed
```

Every graphics device has its own graphics state and calls to `par()`
only affect the traditional graphics state of the currently active
graphics device. 

## 3.2 Controlling the appearance of plots

This section is concerned with the appearance of plots, which means
the colors, line types, fonts and so on that are used to draw a plot. 

`col` controls the color of output. This can be set permanently using
`par()` with an expression of the form `par(col="red")`, which will
affect all subsequent graphical output. `plot(..., col="red")` will
affect the output just for that plot. 

The setting can be used as an argument to a low-level function, as in
the expression below `lines(..., col="red")`, which demonstrates that
the setting can be used to control the appearance of just a single
piece of graphical output. 

### 3.2.1 Colors

Three main color settings in the traditional graphics state: `col`, `fg`, and `bg`.

The `col` is used to specify the color of data symbols, lines, text,
and so on that are drawn in the plot region. Unfortunately, when
specified via a graphics function, the effect can vary. For example, a
standard scatterplot produced by the `plot()` function will use `col`
for coloring data symbols and lines, but the `barplot()` function will
use `col` for filling the contents of its bars. 

The effect of `col` on graphical output drawn in the margins also
varies. It does not affect the color of axes and axis labels, but it
does affect the output from the `mtext()` function. 

The `fg` setting is primarily intended for specifying the color of
axes and borders on plots. 

The `bg` setting is primarily intended to specify the color of the
background for traditional graphics output. This color is used to fill
the entire page. When `bg` is specified in a graphics function it can
have a quite different meaning. For example, the `plot()` and
`points()` functions use `bg` to specify the color for the interior of
the data symbols. 

#### Fill patterns

When printing in black and white, it is difficult to make use of
different colors to distinguish between different elements of a
plot. Using different levels of gray can be effective, but another
option is to make use of some sort of fill pattern, such as
cross-hatching. 

Fill patterns and they can only be applied to rectangles and
polygons. It is possible to fill a rectangle or polygon with a set of
lines drawn at a certain angle, with a specific separation between the
lines. A `density` argument controls the separation between the lines
(in terms of lines per inch) and an `angle` argument controls the
angle of the lines. These settings can only be controlled via
arguments to the functions `rect()`, `polygon()`, `hist()`,
`barplot()`, `pie()`, and `legend()`. 

### 3.2.2 Lines

The `lty` setting describes the type of line to draw (e.g., `solid`,
`dashed`, or `dotted`), the `lwd` setting describes the width of
lines, and the `ljoin`, `lend`, and `lmitre` settings control how the
ends and corners in lines are drawn (rounded or pointy). The scope of
these settings again differs depending on the graphics function being
called. 

### 3.2.3 Text

The size of text is controlled via `ps` and `cex`; the font is
controlled via `font` and `family`; the justification of text is
controlled via `adj`; and the angle of rotation is controlled via
`srt`. 

#### Text size

The size of text is ultimately a numerical value specifying the size
of the font in “points.” The font size is controlled by two settings:
`ps` specifies an absolute font size setting (e.g., `ps=9`), and `cex`
specifies a multiplicative modifier (e.g., `cex=1.5`). The final font
size specification is simply `fontsize * cex`. 

When `cex` is specified via `par()`, it affects most text. However,
when `cex` is specified via `plot()`, it only affects the size of data
symbols. There are special settings for controlling the size of text
that is drawn as axis tick labels (`cex.axis`), text that is drawn as
axis labels (`cex.lab`), text in the title (`cex.main`), and text in
the subtitle (`cex.sub`). 

#### Specifying fonts

The `family` setting is a character value giving the name of a
specific font family, such as "Times Roman", or a generic family
style, such as "serif", "sans" (sans-serif), or "mono" (monospaced). 

"The `font` is a numeric value that selects between normal text (1),
bold (2), italic (3), and bold-italic (4). 

#### Justification of text

The `adj` setting is a value from 0 to 1 indicating the horizontal
justification of text strings (0 means left-justified, 1 means
right-justified and a value of 0.5 centers text). In the plot region,
the justification is relative to the (x, y) location at which the text
is being drawn. 

In the figure margins and outer margins, the meaning of the `adj`
setting depends on the `las` setting. When margin text is parallel to
the axis, `adj` specifies both the location and the justification of
the text. 

#### Rotating text

The `srt` setting specifies a rotation angle anti-clockwise from the
positive x-axis, in degrees. This will only affect text drawn in the
plot region (text drawn by the `text()` function. 

In the figure and outer margins, text may only be drawn at angles that
are multiples of 90 degrees, and this angle is controlled by the `las`
setting. 

#### Multi-line text

The spacing between multiple lines of text is controlled by the
`lheight` setting, which is a multiplier applied to the natural height
of a line of text. For example, `lheight=2` specifies double-spaced
text. This setting can only be specified via `par()`. 

### 3.2.4 Data symbols

The data symbol used for plotting points is controlled by the `pch`
setting. This can be an integer value to select one of a fixed set of
data symbols, or a single character. For example, specifying `pch=0`
produces an open square, `pch=1` produces an open circle, and `pch=2`
produces an open triangle, etc. Specifying `pch="#"` means that a hash
character will be plotted at each data location. 

Some of the predefined data symbols (pch between 21 and 25) allow a
fill color separate from the border color, with the `bg` setting
controlling the fill color in these cases. 

The type setting controls how data are represented in a plot. A value
of "p" means that data symbols are drawn at each (x, y) location. The
value "l" means that the (x, y) locations are connected by lines. A
value of "b" means that both data symbols and lines are drawn. The
type setting may also have the value "o", which means that data
symbols are “over-plotted” on lines (with the value "b", the lines
stop short of each data symbol). 

It is also possible to specify the value "h", which means that
vertical lines are drawn from the x-axis to the (x, y) locations (the
appearance is like a barplot with very thin bars). Two further values,
"s" and "S" mean that (x, y) locations are joined in a city-block
fashion with lines going horizontally then vertically (or vertically
then horizontally) between each data location. Finally, the value "n"
means that nothing is drawn at all. 

### 3.2.5  Axes

The traditional graphics system produces axes with sensible labels and
tick marks at sensible locations. The user may have to resort to
drawing the axis explicitly using the `axis()` function. The `lab`
setting in the traditional graphics state is used to control the
number of tick marks on the axes. 

See the text for additional settings.

### 3.2.6 Plotting regions

The traditional graphics system defines several different regions on
the graphics device. How do we to control the size and layout of these
regions using graphics state settings. The size of each margin can be
controlled independently, but R will check whether an overall
specification is consistent. 

#### Outer margins

By default, there are no outer margins on a page. Outer margins can be
specified using the `oma` graphics state setting. This consists of
four values for the four margins in the order (bottom, left, top,
right) and values are interpreted as lines of text. 

#### Figure regions

By default, the figure region is calculated from the settings for the
outer margins and the number of figures on the page. 

#### Figure margins

The figure margins can be controlled using the `mar` state
setting. This consists of four values for the four margins in the
order (bottom, left, top, right) where each value represents a number
of lines of text. 

#### Plot regions

By default, the plot region is calculated from the figure region less
the figure margins. The location and size of the plot region may be
controlled explicitly instead, using the `plt`, `pin`, or `pty`
settings. 

See Figure 3.11 for more details.

### 3.2.7 Clippings

Traditional graphics output is usually clipped to the plot
region. This means that any output that would appear outside the plot
region is not drawn. For example, in the default behavior, data
symbols for (x, y) locations which lie outside the ranges of the axes
are not drawn. 

Traditional graphics functions that draw in the margins clip output to
the current figure region or to the device. 

It can be useful to override the default clipping region. For example,
this is necessary to draw a legend outside the plot region using the
`legend()` function. 

### 3.2.8 Moving to a new plot

The `devAskNewPage()` function can be used to control whether the user
is prompted before the graphics system starts a new page of output. 


























