---
title: 'ggvis: basics layers'
author: "jharner"
date: "March 10, 2015"
output: html_document
---

## Layers

So far, you seen two layer functions: `layer_points()` and `layer_histograms()`. There are many other layers, and they can be roughly categorised into two types:

* Simple, which include primitives like points, lines and rectangles.
* Compound, which combine data transformations with one or more simple layers.

All layer functions use the plural, not the singular. Think the verb, not the noun: I’m going to layer some points onto my plot.

### Simple layers

There are five simple layers:

* Points, `layer_points()`, with properties `x`, `y`, `shape`, `stroke`, `fill`, `strokeOpacity`, `fillOpacity`, and `opacity`.
```{r}
library(dplyr)
library(ggvis)
mtcars %>% ggvis(~wt, ~mpg) %>% layer_points()
```

* Paths and polygons, `layer_paths()`.
```{r}
df <- data.frame(x = 1:10, y = runif(10))
df %>% ggvis(~x, ~y) %>% layer_paths()
```

If you supply a fill, you’ll get a polygon
```{r}
t <- seq(0, 2 * pi, length = 100)
df <- data.frame(x = sin(t), y = cos(t))
df %>% ggvis(~x, ~y) %>% layer_paths(fill := "red")
```

* Filled areas, `layer_ribbons()`. Use properties `y` and `y2` to control the extent of the area.
```{r}
df <- data.frame(x = 1:10, y = runif(10))
df %>% ggvis(~x, ~y) %>% layer_ribbons()
df %>% ggvis(~x, ~y + 0.1, y2 = ~y - 0.1) %>% layer_ribbons()
```

* Rectangles, `layer_rects()`. The location and size of the rectangle is controlled by the `x`, `x2`, `y` and `y2` properties.
```{r}
set.seed(1014)
df <- data.frame(x1 = runif(5), x2 = runif(5), y1 = runif(5), y2 = runif(5))
df %>% ggvis(~x1, ~y1, x2 = ~x2, y2 = ~y2, fillOpacity := 0.1) %>% layer_rects()
```

* Text, `layer_text()`. The text layer has many new options to control the apperance of the text: `text` (the label), `dx` and `dy` (margin in pixels between text and anchor point), `angle` (rotate the text), `font` (font name), `fontSize` (size in pixels), `fontWeight` (e.g. bold or normal), `fontStyle` (e.g. italic or normal.)
```{r}
df <- data.frame(x = 3:1, y = c(1, 3, 2), label = c("a", "b", "c"))
df %>% ggvis(~x, ~y, text := ~label) %>% layer_text()
df %>% ggvis(~x, ~y, text := ~label) %>% layer_text(fontSize := 50)
df %>% ggvis(~x, ~y, text := ~label) %>% layer_text(angle := 45)
```

### Compound layers

The four most common compound layers are:

* `layer_lines()` which automatically orders by the `x` variable:
```{r}
t <- seq(0, 2 * pi, length = 20)
df <- data.frame(x = sin(t), y = cos(t))
df %>% ggvis(~x, ~y) %>% layer_paths()
df %>% ggvis(~x, ~y) %>% layer_lines()
```

`layer_lines()` is equivalent to `arrange()` + `layer_paths()`:

```{r}
df %>% ggvis(~x, ~y) %>% arrange(x) %>% layer_paths()
```

* `layer_histograms()` and `layer_freqpolys()` which allows you to explore the distribution of continuous. Both layers first bin the data with `compute_bin()` then display the results with either rects or lines.
```{r}
mtcars %>% ggvis(~mpg) %>% layer_histograms()
# Or equivalently
binned <- mtcars %>% compute_bin(~mpg) 
binned %>% 
  ggvis(x = ~xmin_, x2 = ~xmax_, y2 = 0, y = ~count_, fill := "black") %>%
  layer_rects()
```

* layer_smooths() fits a smooth model to the data, and displays predictions with a line. It’s used to highlight the trend in noisy data:
```{r}
mtcars %>% ggvis(~wt, ~mpg) %>% layer_smooths()
# Or equivalently
smoothed <- mtcars %>% compute_smooth(mpg ~ wt)
smoothed %>% ggvis(~pred_, ~resp_) %>% layer_paths()
```

You can control the degree of wiggliness with the `span` parameter:
```{r}
span <- input_slider(0.2, 1, value = 0.75)
mtcars %>% ggvis(~wt, ~mpg) %>% layer_smooths(span = span)
```

## Multiple layers

Rich graphics can be created by combining multiple layers on the same plot. This easier to do: just layer on multiple elements:
```{r}
mtcars %>% 
  ggvis(~wt, ~mpg) %>% 
  layer_smooths() %>% 
  layer_points()
```

You could use this approach to add two smoothers with varying degrees of wiggliness:
```{r}
mtcars %>% ggvis(~wt, ~mpg) %>%
  layer_smooths(span = 1) %>%
  layer_smooths(span = 0.3, stroke := "red")
```

## More details

There are other optional components that you can include:

* `scales`, to control the mapping between data and visual properties. These are described in the properties and scales vignette.

* `legends` and `axes` to control the appearance of the guides produced by the scales. See the axes and legends vignette for more details.




